Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-10-20 analysis/paper.qmd
Finish renaming package
Fix citations in docx

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: jar@hum.ku.dk
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 65 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential niches of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  Our models performed well against independent contemporary test data, but 
  their ability to predict the occurrence of specific species at 
  archaeological sites was much more variable, probably reflecting a tendency of 
  the method to underestimate the species' fundamental niche. Nevertheless, the 
  majority of species are predicted to have had more restricted geographic 
  distributions under past climate conditions compared to today. Crop 
  progenitors and several wild food species modelled are modelled to have been 
  concentrated in the Levant and, to a lesser extent, Cyprus and Western 
  Anatolia, suggesting these regions may have served as glacial *refugia*. 
  The average size of species' niche shrunk by an average of c. 25% from the 
  terminal Pleistocene to the Early Holocene, indicating that economically
  significant plants were adapted to cryo-arid conditions and did not, as often
  assumed, initially respond positively to the 'ameliorated' climate of the 
  Holocene.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
  docx:
    toc: true
    number-sections: true
bibliography: 
  - references.bib
  - gbif.bib
execute: 
  echo: false
  cache: true
---

<!--
TODO before publication:
- [ ] Zenodo release
- [ ] Clean up package code
-->

@@

library("BadiaPaleoFloraENM") # this package, install with devtools::install()
```

@@

In general there is middling to low agreement between our hindcast predictions of species niches and observed archaeological occurrences of the same species [@tlb-results-summary].

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: jar@hum.ku.dk
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 65 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential niches of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  Our models performed well against independent contemporary test data, but 
  their ability to predict the occurrence of specific species at 
  archaeological sites was much more variable, probably reflecting a tendency of 
  the method to underestimate the species' fundamental niche. Nevertheless, the 
  majority of species are predicted to have had more restricted geographic 
  distributions under past climate conditions compared to today. Crop 
  progenitors and several wild food species modelled are modelled to have been 
  concentrated in the Levant and, to a lesser extent, Cyprus and Western 
  Anatolia, suggesting these regions may have served as glacial *refugia*. 
  The average size of species' niche shrunk by an average of c. 25% from the 
  terminal Pleistocene to the Early Holocene, indicating that economically
  significant plants were adapted to cryo-arid conditions and did not, as often
  assumed, initially respond positively to the 'ameliorated' climate of the 
  Holocene.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
  docx:
    toc: true
    number-sections: true
    citeproc: true
    csl: jcaa.csl
bibliography: 
  - references.bib
  - gbif.bib
execute: 
  echo: false
  cache: true
---

@@

library("WAsiaPaleoFloraENM") # this package, install with devtools::install()
```

@@

In general there is middling to low agreement between our hindcast predictions of species niches and observed archaeological occurrences of the same species [@tbl-results-summary].
